---
title: "Kangaroo distance sampling survey along roads"
author: "Marques and Rexstad"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kangaroo distance sampling survey along roads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Some set of introductory comments about the vignette

## Make package functions available

```{r}
library(densgrad)
```

## Acquire data

Data come from three sources:  

- distance sampling survey (transect lengths, perpendicular distances and group size)
- random locations within truncation distance of the transects (road)
    - actually, random data don't appear to be part of the actual analysis
- GPS collar data from telemetered kangaroos within truncation distance of transects

```{r}
real <- read.csv(file="../data/GPSpoints.txt", header=TRUE, sep="\t")
```

Make some adjustments to field names

```{r}
names(real) <- c("infid","nearfid","dist","reserve","veg","animal")
```

Actual distance sampling survey data

```{r}
survey <- read.csv(file="../data/nonstratified_farrar.txt", header=TRUE, sep="\t")
```


### Data munging

Transect 23 had no kangaroo detections, hence in the data the perpendicular distance is recorded as `NA`.  That record needs to be removed from the data set.  Similarly, the truncation distance is set here and the survey data edited accordingly.

```{r}
survey.truncation <- 150
gradient.max.distance <- 100
survey.data <- survey$Distance[!is.na(survey$Distance) & survey$Distance<survey.truncation]
```

## Density gradient estimation

Apparently this optimisation is not really necessary, but perhaps used for plotting?

```{r}
gradient <- optim(par=c(35,1.2),
                  fn=likdensity,
                  distances=real$dist,
                  truncation=gradient.max.distance,
                  control=list(fnscale=-1))#,method="L-BFGS-B",lower=c(20,-100),upper=c(200,100))
if (gradient$convergence == 0) {
  sigma.est <- gradient$par[1]
  beta.est <- gradient$par[2]
  likelihood <- gradient$value
} else {
  print("Convergence not achieved, no parameter estimates produced")
} 
```

Let's see a plot of the fitted animal distribution with respect to transects

```{r, fig.cap="Fitted animal density gradient"}
xs <- seq(from=0, to=gradient.max.distance)
par(mfrow=c(1,1),mar=c(4,4,0.5,0.5))
hist(real$dist,prob=T,main="",xlab="Distance (m)")
lines(xs,pi.x(xs,sigma.est,beta.est,gradient.max.distance),type="l",lty=2,lwd=2)
```

##  Joint likelihood estimation

First the half normal detection function

```{r}
hn.det <- optim(par=c(50,35,35),
                fn=full.lik.fx.HN,
                detectdists=survey.data,
                gpsdists=real$dist,
                truncdet=survey.truncation,
                truncden=gradient.max.distance,
                control=list(fnscale=-1))#,method="L-BFGS-B",lower=c(20,-100),upper=c(200,100))
if (hn.det$convergence == 0) {
  sigma.detect.est <- hn.det$par[1]
  sigma.gradient.est <- hn.det$par[2]
  beta.est <- hn.det$par[3]
  likelihood.hn <- hn.det$value
} else {
  print("Convergence not achieved, no parameter estimates produced")
} 
```

Next the hazard rate detection function

```{r}
hz.det <- optim(par=c(50,1.2,35,35),
                fn=full.lik.fx.HR,
                detectdists=survey.data,
                gpsdists=real$dist,
                truncdet=survey.truncation,
                truncden=gradient.max.distance,
                control=list(fnscale=-1))#,method="L-BFGS-B",lower=c(20,-100),upper=c(200,100))
if (hz.det$convergence == 0) {
  sigma.detect.est <- hz.det$par[1]
  beta.detect.est <- hz.det$par[2]
  sigma.gradient.est <- hz.det$par[3]
  beta.gradient.est <- hz.det$par[4]
  likelihood.hz <- hz.det$value
} else {
  print("Convergence not achieved, no parameter estimates produced")
} 
```

The joint likelihood for the half normal detection model is `r round(likelihood.hn,1)` while the corresponding value for the hazard rate detection model is `r round(likelihood.hz,1)`.

## Plots of estimated hazard rate detection function

```{r, fig.cap="Fitted hazard rate detection function accounting for non-uniform distribution."}
hist(survey$Distance[survey$Distance<survey.truncation],
     main="",yaxt="n",ylab="g(y)",xlab="Distance (m)")
xs <- seq(from=0, to=survey.truncation)
lines(xs,60*detfn(z=xs,pars=hz.det$par[1:2],key="HR",adjn=0,w=survey.truncation),
      ylab="g(y)",lwd=2,type="l",lty=3)
axis(2,at=seq(0,60,length=5),labels=seq(0,1,length=5))
```
